---
title: "Price of Eggs"
format: 
  html:
    toc: true
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


```{r message = FALSE}
library(tidyverse)  # ecosystem of data science packages
library(lubridate)  # for working with dates
```



# Source: NYT

The data visualization of this report is based on the following article 
(by Ashley Wu) of The New York Times (Feb 3, 2023):

__Why Eggs Cost So Much__

<https://www.nytimes.com/2023/02/03/briefing/why-eggs-cost-so-much.html>


::: {.callout-note icon=false}
## Note

We are assuming that you have read the article cited above.
:::



# Data

The data behind the NYT article comes from one of the databases of the 
_US Bureau of Labor Statistics_ (BLS), namely, the
_Consumer Price Index Average Price Data_.

Specifically, the time series data is in the following link: 
<https://data.bls.gov/timeseries/APU0000708111>

- __Series Id__:	APU0000708111
- __Series Title__:	Eggs, grade A, large, per doz. in U.S. city average, average price, not seasonally adjusted
- __Area__:	U.S. city average
- __Item__:	Eggs, grade A, large, per doz.


## Data File

The data is in the text file `"price-of-eggs-timeseries.txt"` available
in the following link:

[price-of-eggs-timeseries.txt](price-of-eggs-timeseries.txt)


Assuming that you have this file in your working directory, you can import it as follows:

```{r}
tbl = read_csv(
  file = "price-of-eggs-timeseries.txt", 
  col_types = cols(
    Year = col_double(),
    Jan = col_double(),
    Feb = col_double(),
    Mar = col_double(),
    Apr = col_double(),
    May = col_double(),
    Jun = col_double(),
    Jul = col_double(),
    Aug = col_double(),
    Sep = col_double(),
    Oct = col_double(),
    Nov = col_double(),
    Dec = col_double())
)

head(tbl)
```


## Data Preparation

For `ggplot` convenience, it's better if we pivot the table to __long__ or
__tall__ format so that we only work with 3 variables or columns: 
`Year`, `Month`, and `Price`:

```{r}
# Pivot to longer format
dat = pivot_longer(
  data = tbl, 
  cols = Jan:Dec,      # columns to be pivoted
  names_to = "Month",  # name of column with categorical values
  values_to = "Price") # name of column with numeric values

head(dat)
```

Since we are working with time series data, we also need to add a `Date` column 
with dates:

```{r}
# add date column
dat = dat |>
  mutate(Date = as.Date(paste0(Year, "-", Month, "-01"), "%Y-%b-%d"))

head(dat)
```


-----


# Graph: Timeline

Keep in mind that the timeline of the NYT article goes from Jan-2000 to Dec-2022.
In our case, we have a few more data points (12 months of 2023).


```{r}
# replicated graphic
ggplot(data = dat, aes(x = Date, y = Price)) +
  geom_hline(yintercept = 0:5, color = "gray90") +
  geom_line(color = "orange", linewidth = 0.8)  +
  geom_point(x = as.Date("2023-01-01"), y = 4.823, color = "orange") +
  scale_x_date(breaks = as.Date(paste0(seq(2000, 2024, by = 5), "-01-01"), "%Y-%m-%d")) +
  scale_y_continuous(breaks = 0:5, labels = paste0("$", 0:5)) +
  annotate(geom = "text", x = as.Date("2022-12-31"), y = 5.3, 
           label = "Jan-2023", hjust = 1, color = "gray70", size = 3.5) + 
  annotate(geom = "text", x = as.Date("2023-01-01"), y = 5, 
           label = "$4.823", hjust = 1, fontface = 2, size = 3.5) + 
  theme(axis.line = element_blank(),
        panel.background = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(title = "Monthly average price of a dozen eggs",
       x = "",
       y = "")
```

